---
title: "calculating cape wOBA"
author: "me"
date: "2024-07-07"
output: html_document
---

```{r}
knitr::opts_knit$set(root.dir = "C:/Users/gabe2/Desktop/Cape work")
setwd("C:/Users/gabe2/Desktop/Cape work")
library(tidyverse)

PBP_23 <- read_csv("data/full_pbp_23.csv")%>%
      mutate(Year = 2023)

PBP_21_22 <- read_csv("data/full_pbp.csv") %>%
      mutate(Year = format(as.Date(Date, format = "%m/%d/%Y"), "%Y")) %>%
      select(-Date) %>%
      filter(Year > 2020)
```

```{r}
PBP_23 <- PBP_23 %>%
      select("GameID", "Inning", "Top_Btm", "Outs", "BatterName",
             "PAResult", "RunningEvent", "BaseOutEvent", "Description",
             "Runner1B", "Runner2B", "Runner3B", 
             "InningPACount", "PitchNumPA", "TeamRuns", "RunsScoredOnPlay", "Year")

PBP_21_22 <- PBP_21_22 %>%
      select("GameID", "Inning", "Top_Btm", "Outs", "BatterName",
             "PAResult", "RunningEvent", "BaseOutEvent", "Description",
             "Runner1B", "Runner2B", "Runner3B", 
             "InningPACount", "PitchNumPA", "TeamRuns", "RunsScoredOnPlay", "Year")

PBP <- rbind(PBP_21_22, PBP_23)

rm(PBP_21_22, PBP_23)
```


```{r}

PBP <- PBP %>% 
      drop_na(Outs) %>% # remove the pinch runner
      mutate(
            # convert runners to dummy variables
            Runner1B = case_when(
                  is.na(Runner1B) == F ~ 1,
                  T ~ 0
            ),
            Runner2B = case_when(
                  is.na(Runner2B) == F ~ 1,
                  T ~ 0
            ),
            Runner3B = case_when(
                  is.na(Runner3B) == F ~ 1,
                  T ~ 0
            ),
            # establish game state
            # example game state, 2 outs and bases juiced would be 2111
            GameState = paste0(Outs, Runner1B, Runner2B, Runner3B)
      ) %>%
      filter(is.na(PAResult) == F | RunningEvent == T | BaseOutEvent == T)
```

# Use hahn's loop to create runs rest of inning
```{r}
# Group by half inning
PBP <- PBP %>% 
      mutate(half_inn_id = paste(GameID, Top_Btm, Inning, sep = "-"))

# this df is half innings with outs in them and runs in them
half_innings <- PBP %>%
  arrange(half_inn_id, InningPACount, PitchNumPA) %>%
  mutate(OutsOnPlay = (lead(Outs) - Outs) %% 3) %>%
  group_by(half_inn_id) %>%
  summarize(RunsInInning = sum(RunsScoredOnPlay), OutsInInning = sum(OutsOnPlay))

# Join on half inning to get a runs scored in each inning
join_PBP <- PBP %>%
  inner_join(half_innings, by = "half_inn_id")

# Make sure 3 outs in inning, dummy var at start of inning
PBP_3outs <- join_PBP %>%
  filter(OutsInInning == 3) %>%
  mutate(NewInning = if_else(half_inn_id == lag(half_inn_id, default = "0"), 0, 1))

# Calculate runs already scored and yet to be scored
PBP_3outs %>% 
  group_by(grp = cumsum(NewInning == 1)) %>% 
  mutate(AlreadyScored = cumsum(RunsScoredOnPlay) - RunsScoredOnPlay) %>% 
  ungroup() %>% 
  mutate(YetToScore = RunsInInning - AlreadyScored) -> run_vars
```

# for loop to remove duplicate game states
```{r}
# selected_row <- c()
# 
# for (i in 1:(nrow(run_vars) - 1)){
#       if(
#             !is.na(run_vars$GameState[i]) & !is.na(run_vars$GameState[i+1]) & 
#             run_vars$GameState[i] == run_vars$GameState[i+1] &
#             run_vars$InningPACount[i] == run_vars$InningPACount[i+1]
#             
#             ) 
#             {selected_row <- c(selected_row, i+1)}
# }
# 
# run_vars <- run_vars[-selected_row,]
```

# create RE24
```{r}
#run_vars23 <- run_vars %>% filter(Year == 2023)

# RE24_23 <- run_vars %>%
#       filter(Year == 2023) %>%
#       group_by(GameState) %>%
#       summarize(
#             instances = n(),
#             sum_runs_yet_score = sum(YetToScore, na.rm = T),
#             expected_runs = sum_runs_yet_score / instances
#       )
# 
# RE24_22 <- run_vars %>%
#       filter(Year == 2022) %>%
#       group_by(GameState) %>%
#       summarize(
#             instances = n(),
#             sum_runs_yet_score = sum(YetToScore, na.rm = T),
#             expected_runs = sum_runs_yet_score / instances
#       )
# 
# RE24_21 <- run_vars %>%
#       filter(Year == 2021) %>%
#       group_by(GameState) %>%
#       summarize(
#             instances = n(),
#             sum_runs_yet_score = sum(YetToScore, na.rm = T),
#             expected_runs = sum_runs_yet_score / instances
#       )

full_RE24 <- run_vars %>%
      group_by(GameState) %>%
      summarize(
            instances = n(),
            sum_runs_yet_score = sum(YetToScore, na.rm = T),
            expected_runs = sum_runs_yet_score / instances
      )
```

# now i need to create a woba event column and a delta_run_exp column in order to find the average rv delta by outcome
```{r}
fieldOuts <- c("Field out", "Fielder's Choice", "Field out - ground into double play",
             "Fielder's choice", "Sacrifice fly", "Sacrifice fly - reach on error")

punchies <- c("Strikeout", "Strikeout - reach on wild pitch",
              "Strikeout - reach on error", "Strikeout - reach on passed ball", 
              "Strikeout - reach on fielder's choice","Stikeout - reach on passed ball")

PBP_withRE <- run_vars %>%
      mutate(
            wOBA_Event = case_when(
                  PAResult %in% punchies ~ "Punchy",
                  PAResult %in% fieldOuts ~ "FieldOut", # no bunts
                  PAResult %in% c("Walk", "Intentional Walk") ~ "BB", # if youre so good they dont even try that should count
                  PAResult == "Hit by pitch" ~ "HBP",
                  PAResult == "Single" ~ "1B",
                  PAResult == "Double" ~ "2B",
                  PAResult == "Triple" ~ "3B",
                  PAResult == "Home run" ~ "HR",
            )
      ) 

PBP_withRE <- left_join(PBP_withRE, full_RE24, join_by(GameState)) %>%
      select(-instances, -sum_runs_yet_score)

PBP_withRE <- PBP_withRE %>%
      mutate(
            NextValue24 = ifelse(grp == lead(grp), lead(expected_runs), 0),
            delta_run_exp = NextValue24 - expected_runs + RunsScoredOnPlay
      )
```

# here we go baby, now we find mean delta rv, then get rv relative to outs, finally have wOBA RV unsclaed
```{r}
guts <- PBP_withRE %>%
      filter(is.na(wOBA_Event) == F) %>%
      group_by(wOBA_Event) %>%
      summarise(rv = mean(delta_run_exp, na.rm = T)) %>%
      arrange(desc(rv))

# now we get rv relative to outs
guts$rv_relOuts <- guts$rv - min(guts$rv)

PBP_withRE <- left_join(PBP_withRE, guts %>% select(wOBA_Event, rv_relOuts), join_by(wOBA_Event))
```

# lets find league wOBA unscaled and league OBP in order to find wOBA scaling
```{r}
lgOBP <- 
      sum(PBP_withRE$wOBA_Event %in% c("1B", "2B", "3B", "HR", "BB"))/
      (
      sum(PBP_withRE$wOBA_Event %in% c("1B", "2B", "3B", "HR", "BB", "Punchy", "FieldOut")) -
      sum(PBP_withRE$PAResult %in% c("Intentional walk", "Intentional Walk"))    
                  )

lgwOBA_Unscaled <- mean(PBP_withRE$rv_relOuts, na.rm = T)
```

# now get wOBA scaling, which is rv_relOuts * (lgOBP / lgwOBA_unscaled)
```{r}
guts$wOBA_RV <- guts$rv_relOuts * (lgOBP / lgwOBA_Unscaled)
```

# now we can get true wOBA value and find 2023 wOBA
```{r}
PBP_withRE <- left_join(PBP_withRE, guts %>% select(wOBA_Event, wOBA_RV), join_by(wOBA_Event))

wOBA <- PBP_withRE %>%
      group_by(BatterName) %>%
      summarise(
            wOBA = round(mean(wOBA_RV, na.rm = T), 3),
            PA = n()
            ) %>%
      filter(PA >= 40) %>%
      arrange(desc(wOBA))
```

```{r}

```

